---
title: "Assignment 10B"
author: "Leon Lin"
date: "2026-04-16"
output: html_document
---


# Setup

```{r}
#| label: setup
library(jsonlite)  
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(knitr)
```


```{r}
laur_url <- "https://api.nobelprize.org/2.1/laureates?limit=1000&format=json"
laur_raw <- read_json(laur_url, simplifyVector = FALSE)

cat("Laureate records returned:", length(laur_raw$laureates), "\n")
```

## Prize Data

```{r}
prize_url <- "https://api.nobelprize.org/2.1/nobelPrizes?limit=1000&format=json"
prize_raw <- read_json(prize_url, simplifyVector = FALSE)

cat("Prize records returned:", length(prize_raw$nobelPrizes), "\n")
```

```{r}
pluck_chr <- function(x, ...) {
  val <- purrr::pluck(x, ..., .default = NULL)
  if (is.null(val) || length(val) == 0) NA_character_ else as.character(val[[1]])
}

laureates_raw_tbl <- tibble(raw = laur_raw$laureates)

# ── Bio level (one row per laureate) ─────────────────────────────────────────
bio <- laureates_raw_tbl |>
  hoist(raw,
    id          = "id",
    known_name  = list("knownName", "en"),
    full_name   = list("fullName",  "en"),
    org_name    = list("orgName",   "en"),
    gender      = "gender",
    birth_date  = list("birth", "date"),
    birth_ctry  = list("birth", "place", "country",    "en"),
    birth_ctry2 = list("birth", "place", "countryNow", "en"),
    prizes      = "nobelPrizes"
  ) |>
  mutate(
    name          = coalesce(known_name, full_name, org_name),
    gender        = coalesce(gender, "organization"),
    birth_year    = as.integer(str_extract(coalesce(birth_date, ""), "^\\d{4}")),
    birth_country = coalesce(birth_ctry, birth_ctry2)
  ) |>
  select(id, name, gender, birth_year, birth_country, prizes)

# ── Prize level (one row per laureate × prize) ───────────────────────────────
prizes_per_person <- bio |>
  select(id, birth_year, birth_country, gender, prizes) |>
  unnest_longer(prizes) |>
  mutate(
    award_year  = as.integer(pluck_chr(prizes, "awardYear")),
    category    = map_chr(prizes, ~ pluck_chr(.x, "category", "en")),
    aff_country = map_chr(prizes, function(p) {
      affs <- p$affiliations
      if (is.null(affs) || length(affs) == 0) return(NA_character_)
      a <- affs[[1]]
      coalesce(pluck_chr(a, "country", "en"),
               pluck_chr(a, "countryNow", "en"))
    })
  ) |>
  select(id, birth_year, birth_country, gender,
         award_year, category, aff_country)

glimpse(prizes_per_person)
```



```{r}
prize_tbl <- tibble(raw = prize_raw$nobelPrizes) |>
  hoist(raw,
    award_year   = "awardYear",
    category     = list("category", "en"),
    laureates    = "laureates",
    prize_amount = "prizeAmount"
  ) |>
  mutate(
    award_year   = as.integer(award_year),
    prize_amount = as.numeric(prize_amount),
    n_laureates  = map_int(laureates, ~ if (is.null(.x)) 0L else length(.x))
  ) |>
  select(award_year, category, n_laureates, prize_amount)

glimpse(prize_tbl)
```

---

# Question 1 — Which countries have produced the most Nobel laureates by birth?

**Question:** Country of birth is a timestable proxy for where laureates
grew up and were educated. Which nations have historically been the most
productive sources of Nobel-level talent?

```{r}
top_birth <- prizes_per_person |>
  filter(!is.na(birth_country), birth_country != "") |>
  distinct(id, birth_country) |>
  count(birth_country, name = "laureates") |>
  slice_max(laureates, n = 20) |>
  mutate(rank = row_number()) |>
  arrange(desc(laureates))

top_birth |>
  select(rank, birth_country, laureates) |>
  kable(col.names = c("Rank", "Birth Country", "Laureates Born There"))
```

```{r}
top_birth |>
  mutate(birth_country = reorder(birth_country, laureates)) |>
  ggplot(aes(laureates, birth_country, fill = laureates)) +
  geom_col() +
  scale_fill_gradient(low = "#a8c6e8", high = "#1a5c9e", guide = "none") +
  labs(
    title    = "Top 20 Countries by Nobel Laureate Birth",
    subtitle = "One row per unique laureate; multiple prizes counted once",
    x        = "Laureates Born There",
    y        = NULL
  ) +
  theme_minimal(base_size = 13)
```

**Answer:** The United States, Germany, and the United Kingdom lead by
birth country. However, many US-affiliated winners were born abroad and
emigrated — the US is a net *recipient* of scientific talent. Germany's
high birth count reflects its 19th-century educational dominance, while
its post-1933 exodus is the subject of Question 2.

---

# Question 2 — Which country has the most "brain drain"? *(Join across fields)*

**Question:** A laureate born in Country A but affiliated with an
institution in Country B at the time of their award represents a loss of
intellectual capital from Country A. Which birth countries lost the most
laureates this way?

*This requires comparing `birth_country` and `aff_country` within the same
laureate record — a cross-field filter that cannot be answered from either
field alone.*

```{r}
drain <- prizes_per_person |>
  filter(
    !is.na(birth_country), birth_country != "",
    !is.na(aff_country),   aff_country   != "",
    birth_country != aff_country
  ) |>
  distinct(id, birth_country, aff_country)

emigrated <- drain |>
  count(birth_country, name = "emigrated")

total_born <- prizes_per_person |>
  filter(!is.na(birth_country), birth_country != "") |>
  distinct(id, birth_country) |>
  count(birth_country, name = "total_born")

drain_tbl <- emigrated |>
  left_join(total_born, by = "birth_country") |>
  mutate(pct_abroad = round(100 * emigrated / total_born, 1)) |>
  arrange(desc(emigrated))
```

```{r}
drain_tbl |>
  slice_head(n = 15) |>
  kable(col.names = c("Birth Country", "Emigrated Laureates",
                      "Total Born There", "% Who Worked Abroad at Award"))
```

```{r}
drain_tbl |>
  slice_head(n = 15) |>
  mutate(birth_country = reorder(birth_country, emigrated)) |>
  ggplot(aes(emigrated, birth_country, fill = pct_abroad)) +
  geom_col() +
  scale_fill_gradient(low = "#fee08b", high = "#d73027", name = "% Abroad") +
  labs(
    title    = "Nobel Prize Brain Drain: Born There, Awarded Elsewhere",
    subtitle = "Laureates whose prize-time affiliation was outside their birth country",
    x        = "Emigrated Laureates",
    y        = NULL
  ) +
  theme_minimal(base_size = 13)
```

```{r}
drain_tbl |>
  filter(total_born >= 5) |>
  slice_max(pct_abroad, n = 15) |>
  mutate(birth_country = reorder(birth_country, pct_abroad)) |>
  ggplot(aes(pct_abroad, birth_country, fill = pct_abroad)) +
  geom_col() +
  scale_fill_gradient(low = "#fee08b", high = "#d73027", guide = "none") +
  labs(
    title    = "Highest Nobel Brain Drain Rate",
    subtitle = "% of born-there laureates affiliated abroad at award (≥5 born laureates)",
    x        = "% Who Worked Abroad at Time of Award",
    y        = NULL
  ) +
  theme_minimal(base_size = 13)
```

**Answer:** **Germany** has lost the most Nobel laureates in absolute
numbers. A large share were Jewish scientists who fled after 1933 and
received prizes at US and UK universities. Hungary and Poland show the
highest *rates* — over 50% of their born laureates won prizes while
working abroad, reflecting World War II displacement and Cold War
political persecution.

---

# Question 3 — How has the average age at award changed over the decades?

**Question:** Is there a widening gap between scientific discovery and
Nobel recognition? If so, laureates should be older on average in recent
decades than they were a century ago.

```{r}
age_data <- prizes_per_person |>
  filter(!is.na(birth_year), !is.na(award_year),
         gender != "organization") |>
  mutate(
    age_at_award = award_year - birth_year,
    decade       = (award_year %/% 10) * 10
  ) |>
  filter(age_at_award >= 25, age_at_award <= 100)

age_by_decade <- age_data |>
  group_by(decade) |>
  summarise(mean_age = mean(age_at_award), .groups = "drop")

age_by_cat <- age_data |>
  group_by(decade, category) |>
  summarise(mean_age = round(mean(age_at_award), 1),
            n        = n(), .groups = "drop") |>
  filter(n >= 3)
```

```{r}
age_by_decade |>
  ggplot(aes(decade, mean_age)) +
  geom_line(linewidth = 1.2, color = "#2c7bb6") +
  geom_point(size = 3, color = "#2c7bb6") +
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed",
              color = "gray50", fill = "gray85") +
  scale_x_continuous(breaks = seq(1900, 2020, 20)) +
  labs(
    title    = "Rising Age at Nobel Prize Award (All Categories Combined)",
    subtitle = "Age estimated as award year − birth year",
    x        = "Decade of Award",
    y        = "Mean Age at Award"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
ggplot(age_by_cat, aes(decade, mean_age, color = category)) +
  geom_line(linewidth = 1) +
  geom_point(aes(size = n), alpha = 0.8) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  scale_color_brewer(palette = "Set2", name = "Category") +
  scale_x_continuous(breaks = seq(1900, 2020, 20)) +
  labs(
    title    = "Average Age at Award by Category and Decade",
    subtitle = "Point size ∝ number of laureates in that cell",
    x        = "Decade of Award",
    y        = "Mean Age at Award"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

```{r}
age_data |>
  group_by(category) |>
  summarise(
    mean_age   = round(mean(age_at_award), 1),
    median_age = median(age_at_award),
    youngest   = min(age_at_award),
    oldest     = max(age_at_award),
    n          = n()
  ) |>
  arrange(desc(mean_age)) |>
  kable(col.names = c("Category", "Mean Age", "Median",
                      "Youngest", "Oldest", "N"))
```

**Answer:** The average age at award has risen by roughly **12–15 years**
since 1901. In the early decades, science laureates were typically in their
mid-50s; today they are usually in their late 60s. The Economics prize
(established 1969) has the oldest average laureates. Literature and Peace
are more stable — those prizes recognize sustained careers or advocacy
rather than a single, dateable experimental discovery.

---

# Question 4 — How often are prizes shared, and is collaboration increasing over time?

**Question:** Nobel science prizes can be shared among up to three people.
If modern science is more collaborative, we should see the share of
three-way splits rising over time and solo prizes declining.

```{r}
share_data <- prize_tbl |>
  filter(!is.na(n_laureates), n_laureates > 0) |>
  mutate(
    share_type = case_when(
      n_laureates == 1 ~ "Solo",
      n_laureates == 2 ~ "Split (2)",
      n_laureates == 3 ~ "Split (3)",
      TRUE             ~ "Other"
    ),
    decade = (award_year %/% 10) * 10
  )

share_by_cat <- share_data |>
  count(category, share_type) |>
  group_by(category) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup()
```

```{r}
share_by_cat |>
  mutate(
    share_type = factor(share_type,
                        levels = c("Solo", "Split (2)", "Split (3)", "Other")),
    category   = reorder(category, -pct, function(x) x[1])
  ) |>
  ggplot(aes(category, pct, fill = share_type)) +
  geom_col(position = "stack") +
  scale_fill_manual(
    values = c("Solo"      = "#2166ac",
               "Split (2)" = "#92c5de",
               "Split (3)" = "#f4a582",
               "Other"     = "#d6604d"),
    name   = "Prize Structure"
  ) +
  labs(
    title = "Nobel Prize Sharing Structure by Category",
    x     = NULL,
    y     = "% of Prizes in Category"
  ) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x  = element_text(angle = 20, hjust = 1),
        legend.position = "bottom")
```

```{r}
science_cats <- c("Physics", "Chemistry", "Medicine")

share_data |>
  filter(category %in% science_cats) |>
  group_by(decade, category) |>
  summarise(pct_three = 100 * mean(share_type == "Split (3)"),
            n         = n(), .groups = "drop") |>
  filter(n >= 3) |>
  ggplot(aes(decade, pct_three, color = category)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_brewer(palette = "Set1", name = "Category") +
  scale_x_continuous(breaks = seq(1900, 2020, 20)) +
  labs(
    title    = "Rise of Three-Way Prize Splits in Science",
    subtitle = "% of prizes in each decade split among three laureates",
    x        = "Decade",
    y        = "% Three-Way Splits"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

```{r}
share_data |>
  count(share_type, name = "prizes") |>
  mutate(pct = round(100 * prizes / sum(prizes), 1)) |>
  arrange(share_type) |>
  kable(col.names = c("Prize Structure", "Count", "% of All Prizes"))
```

---

# Summary found

| # | Question | Key Finding |
|---|---|---|
| 1 | Top countries by birth? | USA, Germany, UK but USA is partly built on immigrant laureate talent |
| 2 | Most Nobel brain drain? | Germany (absolute count); Hungary & Poland (highest emigration rate, of abot 50%+) |
| 3 | Is age at award rising? | Yes — around 12 to15 year increase since 1901. Economics oldest, Peace most stable |
| 4 | Are prizes shared more? | Yes — threeway splits dominate modern Science; solo prizes nearly extinct there |

]
